---
title: "Class 1 - R commands 101"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---


## 1. The very basics of R ##

    All operations in R are performed in the console.
    R console is a powerfull calculator...

<br/>

#### **1.1. The very basic mathematical operarions can be performed in R** ####
```{r}
10+5; 10-5
4*6
16/4
2^8
2*5-(2^5)/8
```
<br/>

#### **1.2 Logical operarions are also possible:** ####
(not relevant for now, but this will be important for programming) [upcoming class]

```{r}
2 == 2 # is it equal?
3 > 2 # is it larger?
3 < 1 # is it smaller?
4 != 0 # is it not equal?
```
<br/>

## 2. Creating objects (or variables) ##

    In R, most data is commonly stored in variables
    These objects are saved in R backstage (workspace)
    With RStudio it's possible to see these objects in the Environment section (right side)

<br/>

#### **2.1 Let's create some objects** ####
```{r}
# objects are assigned using the "<-" characters
x <- 1
y <- 2
str1 <- "hello world"

# The "=" character may also be used but it's not recommended for the sake of clarity

x = 45
str1 = "bye world"

# now objects x and str1 were overwritten 
```
<br/>

```{r}
# to see what's inside an object, simply type the name

x; y; str1

```

<br/>

#### **2.2. Mathematical operations are also applied to objects** ####
```{r}
z <- x+y
z
z <- z + 1
z
```
<br/>

We have now covered four important components in R:

      - numeric data: 1, 45, 667
      - character data: "hello world", "22"
      - objects: x, str1
      - operators: +, -, <, >, !=, ==, <-
    
Note:
    
    Object names can be given using letters and numbers, but should never start with numbers
    Other characters allowed are "_" and "." 
    
    Operations between character and numeric objects are not possible
    
```{r}

x <- 23
y <- 45

str1 <- "hello"
str2 <- "34"

x + y

# the following operations are not allowed
# y - str2
# str2 + str1
```

<br/>

## 3. Functions ##

    A function encloses a group of commands build to automate a specific operation
    Costumized functions can be built by the user
    R has a lot of functions already available
    
    Calling a function on the console: myFunction(argument, ...)
    Never add space between function name and parenthesis

<br/>

#### **3.1 Some basic functions** ####
```{r}
#How to get a list of the objects stored in our workspace
ls() 
#or
objects()

```
<br/>

```{r}
# remove an object from the workspace
newObjectToDelete <- "I'm a temporary file"
ls()
rm(newObjectToDelete) 
ls()

```
<br/>
```{r}
#check the class of the object x
class(x)
#check the class of the object str1
class(str1)

#check the class of the and expression
class(2 > 5)
```


<br/>

#### **3.2 Other functions are usefull for mathematical operations:** ####

```{r}
# square root
sqrt(1024) 

# mean 
example_vector <- c(1.333, 1.445, 1.555, 1.431) # the c(..., ...) will be explained afterwards

mean(example_vector)
#... and standard deviation
sd(example_vector)

#exponential
exp(2) #e^2

```

<br/>

#### **3.3 Most built-in functions have options (arguments), often with default values** ####
```{r}
#logarithm
# type ?log in the console to see the manual of log() function

log(4) # default base value is exp(1)
log(4, base = exp(1))

log(4, base = 10)
log10(4)  # log function comes in different flavours, such as log10

# round a decimal number
number2round <- 2.633672

round(number2round, digits = 4)
round(2.633672) # default value for digits is 0
round(2.633672, digits = 0)
```

<br/>

Note:

    To get help for any built-in functions just type in the console:
      - ?round
      - help(round)

<br/>

#### **3.4 How to create a new function** ####

Costum function can be created using the following format:

*myFunction <- function(arg1, arg2, ..., argN) { \n
  < any formula or script that usus the arguments >
  return(value)
}*

Here are some examples:
```{r}
# calculate the standard error of the mean
# in this case, 'x' will be the argument for the function to work on

mySEM <- function(x){
  return(sd(x) / sqrt(length(x)))
}

# after running this commands the function names appear in the workspace environment

mySEM(example_vector) 

```
```{r}
# Another Example:
# fahrenheit_to_celsius converter 

fahrenheit_to_celsius <- function(temp_F) {
  temp_C <- (temp_F - 32) * 5 / 9
  return(temp_C)
}

fahrenheit_to_celsius(102)

# Adding arguments with default values

fahrenheit_to_celsius <- function(temp_F, d = 5) {
  temp_C <- (temp_F - 32) * 5 / 9
  temp_C <- round(temp_C, digits = d)
  return(temp_C)
}

fahrenheit_to_celsius(100)

# 'd' is a positional argument with a default value. 
# If another value is given in the second position, this will replace the default
 
fahrenheit_to_celsius(100, 3)

# for clarity, it is better to call arguments by their name
fahrenheit_to_celsius(100, d = 3)
```

<br/>

#### **3.5 Other relevant functions** ####

```{r}
getwd() # get working directory
dir <- getwd() 
dir

# to set a new working directory use setwd(/path/to/newDirectory/)
```

<br/> 

## 4. Data structures: Vectors and Data frames

    Numeric and/or character data can be stored in structured objects
    These include vectors, matrices and data frames, among others

<br/>

#### **4.1 Vectors** ####

    Vectors are homogeneous data structures, organized in 1 dimension
    
<br/>

```{r}

x <- c(34, 35, 33, 34) # Combine Values into a Vector or List
```


```{r}
mean(x)
min(x)
max(x)
log10(x)
```


```{r}
s <- c( min(x),max(x) )
s
```

<br/>



```{r}
x <- c(34, 35, 33, 34)
y <- c("ITQB", "IGC", "IMM")
w <- c("ITQB", 10, "IGC", 34)

class(w)
class(x)
class(y)

# a mixed vector will automatically convert all elements to character
w
```

<br/>

**4.1.2 Operations on vectors**


```{r}
# creating some vectors
v1 <- c(6,7,8)
v2 <- c(3,4,5)
c1 <- c( "itqb", "igc", "imm", 3)
```


*for numeric vectors*
```{r}
# multipy all elements by a constant
v1 * 3 

# operation may be also applied to vectors of the same length 
v1 + v2

```

*for character vectors*

```{r}
toupper( c1 )
```

<br/>

#### **4.2 Data Frames** #### 


    A data frame is a heterogeneous data structure, organized in 2 dimensions 
    
    It's used to store tables, particularly if they have mixed  data types (e.g.  character and numeric).

    Another common format to store data is a *matrix*, which should be only used to store data in a single form (character or numeric).  

    If these conditions are guarenteed one can change the format of a given table from dataframe to matrix (or vice-versa) using these functions: as.data.frame(myMatrix) or as.matrix(myTable)

<br/>

**4.2.1 Tables can be imported from files, but you can make your own tables directly in R**
```{r}
var1 <- c( "John", "Paul", "George", "Ringo", "Geri", "Mel B", "Mel C", "Emma", "Victoria")
var2 <- c( 123456, 234567, 345678, 456789, 344536, 223443, 554345, 234674, 234564)
var3 <- c( 8.5, 9.0, 5.0, 6.75, 9.9, 9.6, 5.3, 7.75, 5.8)
var4 <- c( "Male", "Male", "Male", "Male", "Female", "Female", "Female", "Female", "Female")
#use function data.frame() to create a new table
classroom <- data.frame( ID=var2, Score=var3, Gender=var4 ) 

#use function colnames() and rownames() to retrieve column and row names respectively
colnames(classroom)
rownames(classroom)

# rownames() may also be used to define new names for the rows
rownames( classroom ) <- var1
rownames(classroom)

# check the table by printing only some lines
head(classroom) # first six lines
tail(classroom) # last six lines
```


<br/>

#### **4.3 Getting Data from Vectors and Data Frames** #### 

To explore a vector or dataframe (or matrix), use square brackets and add indexes inside
```{r}
# to get a given value from a vector or dataframe, indexes can be used 

# what's the second value of vector c1?
c1[2]
c1[c(2,3)]
```


```{r}
classroom[1]
# in data frames you can explore the 2 dimensions
# usinf the square brackets and the comma [ rowIndex , columnIndex ]
classroom[,1]
classroom[1,]
classroom[1,2]
```


```{r}
#instead of using indexes, you may also use column and/or rownames
classroom["Paul","Score"]

# to get all the columns of a given line, we leave empty the side after the comma
classroom["Paul", ]

```


```{r}

# an alternative to get values based on the column name usesthe $ symbol ... myTable$columnA
# to get all values from column ID
classroom$ID 

# why is this relevant? 
# this is a good way to filter or access data according to the content of the cells
# e.g. 
#to access the all the elements of a given line, for ID 234567
 
classroom[classroom$ID == 234567, ]

# get a new table for students with score higher than 7
good_classroom <- classroom[classroom$Score >= 7, ]
good_classroom

# get a new table just with ID and Score
classroom_short <- classroom[ , c(1,2)]
classroom_short 
# or
# classroom_short <- classroom[ , c("ID","Score")]
```

### Activity 1

**a) Functions and vectors:**

- a.1) What is the result of the following commands (try to guess without running the code):
    - var1[c(1,4)]
    - var1[1,4]
    - var1[5:9]
    - var1[5:length(var1)]
    - var1[5:(length(var1)-1)]
    - var1[5:length(var1)-1]
    - var1[order(var1)]

- a.2) Write a function called edges() that takes a vector of variable size and returns a vector with the first and the last elements. 

```{r}
# Some help:
# If the variable v refers to a vector, then v[1] is the vector’s first element and v[length(v)] is its last (the function length returns the number of elements in a vector). 

```


<br/>

**b) Dataframes:**
The following command will create a data frame with randomized data:
```{r}
RECORDS <- data.frame(
ID=seq(from=1, to=40, by=1),
Gender=c(rep("Male",times=20),rep("Female",times=20)), 
Age=round(rnorm(40, mean=55, sd=5),digits=0),
Weight=c(rnorm(20, mean=70, sd=10),rnorm(20, mean=70, sd=15))
)
```

- b.1) There are many functions used in this command, 5 to be more exact.
    - Can you identify them and discover what they do?
- b.2) Take a look at the first 10 rows of RECORDS 
    - *hint: type ?head and check the arguments taken by the function* 
- b.3) Get a new table just for women and another just with columns Age and Gender
- b.4) Select only the individuals with score higher than the mean score 
- b.5) Select only the men with a score higher than 70 and age lower than 55 (Google "using and or in R")
- b.6) Order/sort table by weight (Google it)

--------
