---
title: "Class 5 - Programming in R"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

<br/>

## 1. Conditional expressions ##

<br/>

When building specific functions or a script, some conditions may be explicitly stated to allow specific operations.

#### **1.1. "if" statement ** ####

<br/>

Generically, this expression is written this form:

    if (test_expression) {
      statement
    }

```{r}
#example

x <- 5
if(x > 0){
print("Positive number")
}

# Check if a numeric object is even.
# [The remainder of an even number divided by 2 is 0]
# [operator %% outputs the remainder of a division]

y <- 6

if(y %% 2 == 0){
print("This number is even")
}

if(x %% 2 == 0){
print("This number is even")
}

```

<br/>

#### **1.2. "if/else" statement ** ####

<br/>

The presence of the "if", conditions the script to a specific path. Somtimes there's also a possibility to add one or multiple conditional options. 
    
    if (test_expression) {
      statement1
    } else {
      statement2
    }
    
    - "if this condition is met, go this way, else got the other way"
    
```{r}
#if…else statement

x <- -5
if(x > 0){
  print("Non-negative number")
  } else {
  print("Negative number")
}

```

<br/>

There is also the possibility of adding multiple alternative pathways

    if ( test_expression1) {
      statement1
    } else if ( test_expression2) {
      statement2
    } else if ( test_expression3) {
      statement3
    } else {
      statement4
    }


Yet, in this case all these conditions need to be mutually exclusive since only one statement will get executed depending upon the test_expressions.

```{r}
x <- 0

if (x < 0) {
  print("Negative number")
} else if (x > 0) {
  print("Positive number")
} else
  print("Zero")

```

--------

<br/>

#### Activity 5.1 #####

- a.1) Complete the function "sum_eval()", that takes a numeric (integer) vector of any length, sums all the elements and prints "Sum is even" or "Sum is odd".

```{r}
# sum_eval <- function(arg){
#   sum.res <- sum(arg)
#   if ( _________ ) { __________ }  else { ________ }
# } 

#### test the function
# vector_a <- c( 34, 56, 25,64,51, 55, 89)
# vector_b <- c( 78, 43, 90, 64, 3, 34, 89)

# sum_eval(vector_a)
```

- a.2) Complete the function "itqb_search()", that takes a vector of words, of any length, and prints "itqb is present" if "itqb" is one of the words present, or "no hit" if not present. *hint:function tolower converts all character srings to lower case*

```{r}

# itqb_search <- function( ___ ){
#   if ( _____ %in% tolower( arg ) ) { desision <- _____ }  else { decision <- ____ } 
# return(decision)
# }

## test the function
# vector_c <- c("Champalimaud", "IGC", "IMM", "IBMC", "CIBIO")
# vector_d <- c("ITQB", "open", "day")
# itqb_search(vector_c) 
```

- a.3) Write a function that takes two arguments, a numeric p-value and a significance value, and evaluates if H0 should be rejected, using alpha of 0.05.

<br/>

## 2. for loops ##

Loops are used in programming to repeat a specific block of code, a specific amount of times

A for loop is usually used to iterate over a vector.


    for i in vector {
    ... statement ...
    }


if the vector has 5 elements, a cycle will be created and the statement will be run 5 times. At every cycle, i will iterate over the elements of the vector

```{r}

random_vector <- c(45,3333333,345,1,0)

for (i in random_vector) { 
  print(i)
  print("end of cycle")
}
```



```{r}
print("new vector")
name_vector <- c("Ana","Joao","Miguel")

for (i in name_vector) { 
  print(i)
  print("end of cycle")
}

```

We can set a counter inside the loop to count how may cycles were run (length of the vector)

```{r}
n = 0
for (i in random_vector) { 
  n = n + 1
  print("couting")
}
n
```


If statements and for loops can be used in the same operations

```{r}
# screen a vector and count the number of numbers higher than 5
random_vector <-c(3.4, 5.5, 4.9, 5.6, 6.6, 7.8, 1.3, 6.7)
count <- 0

for (i in random_vector) {
  if (i > 4) {
    count <- count + 1 
  }
}
count

```

--------

<br/>

#### Activity 5.2 #####

a.4) complete the following function that takes a numeric vector a counts the number of even numbers inside.

```{r}
# myEven <- function( v ) {
#   count = ____
#   for (number in _____ ) {
#     if ( _____ ) { count = count +1 }
#   }
#   return(count)
# }

# vector_a <- c(2,3,6,5,4,56,67,86)
#myEven(vector_a)

```

a.5) write a function that takes a vector of numbers higher and lower than 0 and returns another vector only with positive values. 

*hint: you need to create an empty vector before starting the loop*

*hint: to include 34 in a vector: vector <- c(vector, 34)*

```{r}
myPositive <- function(v){
  positive <- c()
  for (element in v) {
    if (element > 0){
      positive <- c(positive, element)
    }
  }
return(positive)
}

test_vector <- c(-1,2, -3, 5, 6, -15, 56)
myPositive(test_vector)
```
<br/>

## 3. Automated analysis ##
<br/>

Things to think before starting build a script to analysi multiple data:
- files to analyse must be structured 
- names of files should be also structured (if they contain information in the ID of the samples is a plus)
- define the output : in this case, iteratively import csv files to extract the only the first column, sample name and condition, and add this information in a final table that merges the information from all files. The final table should contain three columns: <sample><condition><value>  

<br/>

using list.file() we can list the files of interest (that match a pattern) and save these names in an object
```{r}

files <- list.files(path = "./", pattern = "201808")

# Now we will create a function to iterate over all files and collect the informaton

makeTable <- function(files) {
  
  #first we open an empty table to host the values of interest
  table <- data.frame(sample = factor(),  
                          condition = factor(), ETR = numeric())
  # then we start the loop
  for (file in files){
    # open a file
    df1 <- read.csv( file, row.names = 1 )
  
  # decompose the name of the file to remove IDs
  # strsplit() separates the name of the file by "_"
  # unlist convert the result of strsplit() into a vector to facilitate indexing
    file_data <- unlist(strsplit(file, "_"))
    # Now we would like to change Control to Mock
    if (file_data[1] == "Control") { file_data[1] <- "Mock" }
  # collect all ETR measurements, they are always in column 1
    ETR_m <- df1[,1]
  # create a vector with the factor Control/Cold contained in file, 
  # with the same length as the ETR measurements
    condition <- rep(file_data[1], length(ETR_m))
  # do the same with line
    sample <- rep(file_data[2], length(ETR_m))
  # create a temporary dataframe by binding all three vectors by column
    temp_df <- cbind.data.frame(sample , condition, ETR_m)
  # bind temporary dataframe to the master_df
  # we will bind by rows
    table <- rbind(table, temp_df)
  }
  return(table)
}  

ETR_table<- makeTable(files)

#create a new column with a new factor resulting from the combination of both factors

ETR_table$samplecondition <- interaction(ETR_table$sample, ETR_table$condition)

boxplot(formula = ETR_table$ETR_m ~ ETR_table$samplecondition)


```

In this case we have two factors (condition, sample) for one continuous variable (ETR_m), so we can compute a Two-way ANOVA. For the sake of brevity, let's assume that samples follow a normal distribution and have equal variances.

*rule of thumb: ANOVA is robust to heterogeneity of variance so long as the largest variance is not more than 4 times the smallest variance*

```{r}
res.aov <- aov(ETR_m ~ condition + sample, data = ETR_table)
summary(res.aov)
TukeyHSD(res.aov)

```

```{r}

res.aov <- aov(ETR_m ~ condition + sample + condition * sample, data = ETR_table)
summary(res.aov)
TukeyHSD(res.aov)

```



## 3. EXTRA: Building PCA for multivariate analysys ##

```{r}
#install.packages("ggplot2")
#install.packages("ggfortify")

library(ggplot2)
library(ggfortify)

df <- iris[c(1, 2, 3, 4)]

autoplot(prcomp(df))

autoplot(prcomp(df), data = iris, colour = 'Species',
         loadings = TRUE)

autoplot(prcomp(df), data = iris, colour = 'Species',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)


```

